Dissecting the relationship between protein structure and sequence variation

### 2015-05-18 manuscript/voronoi/mnv.tex
MN update

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Dissecting the relationship between protein structure and sequence variation}
\vfill
\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}
    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural characteristics have been proposed over the past decade to answer this question. Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability of proteins. A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which by its definition represents the combined effects of local packing density and long-range amino acid interactions on sequence variability. Here we propose a methodology to segregate and quantify the extent to which long range amino acid interactions determine the general patterns of sequence variability in proteins, independently of local packing density. We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse of the volume of the Voronoi cells representing individual sites in protein.

@@

    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}. \\

    Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently. It is perceivable that quantities such as B factor, RSA, and CN, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility. \citet{franzosa_structural_2009} used a variety of the local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral impact. Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in shaping the observed sequence variability patterns of enzymatic proteins in contrast to RSA and local flexibility measures. \\

    It is notable that the site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein. A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure. An estimate of the accessible volume for each site in protein can be generally obtained through a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}. In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009}. Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN. A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $7\AA$ to $13\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009}. \\

    In an attempt to provide a more general definition of CN, some studies \citep[e.g.,][]{lin_deriving_2008} have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, $WCN_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,

        \begin{equation}
            \label{eqn:wcn_pwrl}
            WCN_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}},
        \end{equation}

    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel, which is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}. Moreover, no physical model has been so far proposed to support the power-law kernel used in the definition of WCN and the specific value of exponent often used. \\

    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, their definitions does not involve any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution. This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of local packing density and site-specific flexibility. Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number (CN) and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level. \\

    Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution. Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility. \\

@@

    In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. The {\it ddG rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}. In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids. Therefore, a low ddG rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales. \\

    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins can be defined and obtained following the stability threshold model of Bloom et al. (2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al. (2014) \cite{echave_relationship_2014}. Based upon this model which was extensively described in Chapter \ref{chap:sschar}, a quantity $\Delta\Delta G$ {\it rate} (or {it ddG rate}) was derived for each individual site in all proteins in dataset. A high ddG rate for the $i^{th}$ site in a protein indicates a high stability of the site and the overall conformation of the protein to perturbations caused by substitution of the amino acid residing the site. \\

    As a measure of local flexibility or fluctuation in different parts of the protein structure the temperature, factors ({\it B factor}) for all atoms in PDB files were extracted (c.f., Chapter \ref{chap:sschar}). Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.  Alternatively, the site-specific fluctuation could be calculated from MD simulations. This was however impossible for this study for the large dataset of $209$ proteins considered in this work. \\

@@

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{fractal_dim_max_extent_volume.pdf}
        \end{center}
        \caption{The scaling behavior of protein maximum extent as defined by Eqn. \ref{eqn:max_extent} with protein volume for $209$ monomeric enzymes in the dataset. The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.47\pm0.06$.}
        \label{fig:maxextent}
    \end{figure}

    \begin{figure}
        \begin{center}
        \includegraphics[width=5.5in]{fractal_dim_rad_gyration_nres.pdf}
        \end{center}
        \caption{The scaling behavior of protein's radius  as defined by Eqn. \ref{eqn:rg} with protein length for $209$ monomeric enzymes in the dataset. The mean \& median length of the proteins are $362$ \& $315$ respectively. The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.60\pm0.08$.}
        \label{fig:rg}
    \end{figure}

>>>>

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Dissecting the relationship between protein structure and sequence variation}
\vfill
\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}
    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural characteristics have been proposed over the past decade to answer this question. Most importantly, the role of local packing density has been highlighted and shown recently to be the dominant factor in shaping the observed patterns of site-specific sequence variability of proteins. A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which represents by definition, the combined effects of local packing density and long-range amino acid interactions on sequence variability. Here we propose a methodology to segregate and quantify the extent to which long-range interactions determine the general patterns of sequence variability in proteins, independently of local packing density. We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of the local packing density, defined as the inverse volume of the Voronoi cell corresponding to each site in protein.

@@

    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}. \\

    Although these structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently. It is perceivable that quantities such as B factor and RSA, all serve as proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids. \citet{franzosa_structural_2009} used a variety of local packing density measures to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence. Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in shaping the observed sequence variability patterns of enzymatic proteins in contrast to RSA and local flexibility measures. \\

    It is notable that the site-specific flexibility is often represented by $C_\alpha$ atomic B factor, a quantity that is not necessarily an unbiased measure of the amino acid flexibility as a whole in a given site in protein. A more accurate measure of amino acid flexibility requires the calculation of accessible free volume to each site in protein structure. An estimate of the accessible volume to each site in protein can be generally obtained through the inverse of a quantity widely known as Contact Number introduced and discussed by several authors \citep[e.g.,][]{liao_protein_2005}. In its simplest mathematical form, the Contact Number for a given site in protein is defined as the number of amino acids within a fixed radius $r$ of neighborhood around it \citep[e.g.,][]{franzosa_structural_2009}. Individual sites are generally represented by the coordinates of $C_\alpha$ backbone atoms for the calculation of CN. A major problem with the traditional definition of contact number however, is the existence of the arbitrary parameter $r$ in the definition of CN. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range $5\AA$ to $18\AA$ \citep[e.g.,][]{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014}. \\

    In an attempt to provide a more general definition of CN, some studies \citep[e.g.,][]{lin_deriving_2008} have already suggested an alternative definition known as the Weighted Contact Number (WCN): For a given site $i$ in a protein of length $N$, WCN$_i$ is defined as the sum of the inverse-squared of distances between the amino acid of interest and all other sites in protein,

        \begin{equation}
            \label{eqn:wcn_pwrl}
            \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{2}}.
        \end{equation}

    Although WCN is in general a better predictor of $C_\alpha$ atomic B factor and site-specific sequence variability, the proposed definition of WCN still involves an adjustable free parameter, the exponent of the power-law kernel. The value of the exponent that results in the best predictions of B factors appears to be in the range $-3\lesssim\alpha\lesssim-2$ and is typically fixed to $\alpha=-2$ as shown in Eqn \ref{eqn:wcn_pwrl} \citep[e.g.,][]{yang_protein_2009}.  A similar exponent value was also used by \citet{huang_mechanistic_2014} to argue for the {\it local} packing density of sites as the dominant factor in shaping protein sequence variability. The specific value of $\alpha=-2$ however, implies that the long-range amino acid interactions also play a non-negligible role in sequence evolution, in addition to local packing density. Similarly, the best performing cutoff values in the definition of Contact Number are in the range of $10\AA\lesssim\alpha\lesssim15\AA$, also indicative of the significant influence of the long-range amino acid interactions on B factor and sequence variability, aside from local packing density. In this regard, the two commonly-used proxy measures of packing density may not be considered {\it local}. An important question therefore remains unanswered as to what extent the local packing density and long-range interactions influence sequence variability, separately from each other. \\

    Motivated by the existing gaps in the current understanding of the role of flexibility, packing density and long-range interactions on sequence-structure relations in proteins, here we derive a new set of site-specific structural characteristics that, unlike CN and WCN, do not involve free adjustable parameters in their definitions. In particular, we propose a methodology to derive a minimally-biased measure of {\it local} packing density based upon which the role of long-range interactions on sequence variability can be segregated and isolated from the effects of local packing density.  This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of site-specific packing density and flexibility.

    %Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution. Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility. \\

@@

    In addition to site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.

    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins was also defined and obtained following the stability threshold model of Bloom et al. (2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al. (2014) \cite{echave_relationship_2014}. Based upon this model, the \ddg {\it rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}. In brief, the site-specific quantity, \ddg rate (also denoted by {\it ddG rate} hereafter in figures), is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids. Therefore, a low \ddg rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales. \\

    As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted. Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins. Alternatively, the site-specific fluctuation could be calculated from MD simulations. This was however impossible for this study for the large dataset of $209$ proteins considered in this work. \\

@@

%    \begin{figure}
%        \begin{center}
%        \includegraphics[width=5.5in]{fractal_dim_max_extent_volume.pdf}
%        \end{center}
%        \caption{The scaling behavior of protein maximum extent as defined by Eqn. \ref{eqn:max_extent} with protein volume for $209$ monomeric enzymes in the dataset. The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.47\pm0.06$.}
%        \label{fig:maxextent}
%    \end{figure}
%
%    \begin{figure}
%        \begin{center}
%        \includegraphics[width=5.5in]{fractal_dim_rad_gyration_nres.pdf}
%        \end{center}
%        \caption{The scaling behavior of protein's radius  as defined by Eqn. \ref{eqn:rg} with protein length for $209$ monomeric enzymes in the dataset. The mean \& median length of the proteins are $362$ \& $315$ respectively. The red line is the linear Deming regression fit to logarithms of the two variables with a slope of $D\simeq2.60\pm0.08$.}
%        \label{fig:rg}
%    \end{figure}
